### 2020-08-22 latex/sections/background.tex
latexing

In this chapter we discuss the algorithms we decided to use, what some other similar 
methods exist as well as the field in general.

@@

Categorizing Breakthrough with the characteristics described in Section \ref{sec:environments}. We end up with the description 
shown in Table \ref{tab:breakthrough_cat}. This description is identical to that of Tic-Tac-Toe, and Chess.
This description is the most common in board games where two player compete.

@@

A different heuristic would be the distance of your most advanced pawn minus your opponents most 
advanced pawn, this heuristic could give you insight into how close you are to winning the game or 
how close your opponent is. As a higher level concept we can call this concept your aggressiveness, 
as it closely resembles how aggressive you are for going in for the win. Generally in Breakthrough 
it is favorable to move your whole team as a unit and play more defensively, so this heuristic 
is probably not one that will lead to the most wins.

@@

Traditionally methods for playing games were searching through the environment using a heuristic to
guide the search. A simple way of doing a heuristic based search would be to give all states a $0$ 
as a score and give states where you win a positive score. We say that those search algorithm 
is not guided, and the algorithm will probably have to evaluate every state in the state-space. 
This method of searching is generally extremely inefficient as the state-spaces of game environments 
are generally extremely large, and often infinite. For instance an estimate of the state-space 
for the game breakthrough is $M * N * 2^{N*4}$ where $M$ is the height of the board $N$ is the width 
of the board. The $2^{N*4}$ represent each piece either being alive or captured. So for a small board
$5x4$ the estimated state-space is $1310720$ many states.

@@

The algorithms that are used in traditional state-space search are for instance Depth-first search (DFS),
Breadth-first Search (BFS), Alpha* Search (A-Star), and Monte-Carlo Tree Search (MCTS).

>>>>

In this chapter we discuss traditional artificial intelligence in the context of search the algorithms we used, what some other similar 
methods exist as well as the field in general.

@@

Categorizing Breakthrough with the characteristics described in Section \ref{sec:environments}. We end up with the description 
shown in Table \ref{tab:breakthrough_cat}. These characteristics are identical to that of Tic-Tac-Toe, and Chess.
This categorization is the most common in board games where two player compete.

@@

A different heuristic would be the distance of your most advanced pawn minus your opponents most 
advanced pawn, this heuristic could give you insight into how close you are to winning the game or 
how close your opponent is. As a higher level concept we can call this concept your aggressiveness, 
as it closely resembles how aggressive you are for going in for the win. Generally in Breakthrough 
it is favorable to move your whole team as a unit and play more defensively (CITE UNITY HEURISTIC), so this heuristic 
is probably not one that will lead to the most wins.

@@

Traditionally methods for playing games were searching through the environment using a heuristic to
guide the search. A simple way of doing a heuristic based search would be to give all states a $0$ 
as a score and give states where you win a positive score. We say that those search algorithm 
is not guided, and the algorithm will probably have to evaluate every state in the state-space. 
This method of searching is generally extremely inefficient as the state-spaces of game environments 
are often extremely large, and sometimes infinite. For instance an estimate of the state-space 
for the game breakthrough is $M * N * 2^{N*4}$ where $M$ is the height of the board $N$ is the width 
of the board. The $2^{N*4}$ represent each piece either being alive or captured. So for a small board
$5x4$ the estimated state-space is $1310720$ many states.

@@

The algorithms that are used in traditional state-space search are for instance Depth-first search (DFS),
Breadth-first Search (BFS), Alpha-Beta Pruning Search (AB-Search), and Monte-Carlo Tree Search (MCTS).

### 2020-08-22 latex/sections/introduction.tex
latexing

DL in conjunction with a randomized State-space Search Algorithm Monte-Carlo Tree Search, 
was used to defeat the standing world champion Lee Sedol in the incredibly expansive game of Go. This was done in the year 
2017, the researchers used reinforcement learning (RL), where the agent learned by only playing against itself. 
These result imply that given enough time to learn the computer agent is able to
gain a deep insight on how the game should be played.

>>>>

DL in conjunction with a randomized State-space Search Algorithm Monte-Carlo Tree Search, 
was used to defeat the standing world champion Lee Sedol (CITE ALPHAGO ZERO) in the incredibly expansive game of Go. This was done in the year 
2017, the researchers used reinforcement learning (RL), where the agent learned by only playing against itself. 
These result imply that given enough time to learn the computer agent is able to
gain a deep insight on how the game should be played.

### 2020-08-22 latex/sections/methods.tex
latexing

In this chapter we train a neural network to play the game breakthrough described in 
Chapter \ref{cha:background}. The training process uses MCTS described in Section \ref{sec:mcts} 
to guide it's training.

The training algorithm is a reinforcement learning based self-play algorithm based on the work 
done in AlphaZero (CITE ME).

\subsection{Neural network Architecture}

The neural network architecture we opted to use was a single convolutional layer with a ReLU 
activation function, followed by 5 residual layers, and lastly a split policy/value head.
A figure describing the architecture can be seen in Figure \ref{fig:nnarch}.

@@

\subsection{Policy head}

The policy head of the neural network returns a vector of the size of the action space $w * h * 6$ where $w$ is the width of the board $h$ is the height of the board and $6$ represents the six cardinal direction pawns can move in the space (diagonally both directions, forward and backwards). The vector is then masked s.t. the values representing moves that can not be played on the board are given the value of $0$.

@@

The data collected is $(\pi, \tau, p, v)$, where $\pi$ is the action probabilities provided by 
MCTS for $N$ iterations, $\tau$ is the reward for the whole episode, $1$ if white wins,
$-1$ if black wins, and $0$ on a tied game, $p$ is the policy vector
from the neural network, and $v$ is the predicted reward of the game by the neural network.

@@

>>>>

In this chapter we describe our implementation of the game Breakthrough and the methods for training a neural network to play the game Breakthrough, described in Chapter \ref{cha:background}. The training process uses MCTS described in Section \ref{sec:mcts} to guide its training. We outline the architecture used in the neural network.

The training algorithm is a reinforcement learning self-play algorithm based on previous work by DeepMind (CITE ME).

\section{Implementation}

In this section we will take a brief look at the implementation of the game Breakthrough, how we model its state representation, and the pseudocode for training.

\subsection{Breakthrough}

In order for a neural network to be able to use input, that input needs to be numeric. We model the board in gamethrough with a single $NxM$ array with three values $\{-1,0,1\}$. Where each $x,y$ index on the board represents what value is on the board at the time, $-1$ represnts a white piece, $0$ represents an empty square, and $1$ represents a black piece.

A action vector is coupled with each state, the action vector is a $3$-dimensional matrix where the lengths of the dimensions are $row x columns x 6$ and each index of the vector $X,Y,Z$ represents an action moving from cell $x,y$ moving to the direction $z$. The directions are as follows $0$ represents moving upwards and to the left diagonally on the board, $1$ upwards, and $2$ upwards and to the right diagonally, $4$ downwards and to the left diagonally, $5$ downwards, and lastly $6$ downwards and to the right diagonally.

\subsection{Selfplay}

\section{Neural Network Architecture}

The neural network architecture we opted to use was a single convolutional layer with a ReLU activation function, followed by 5 residual layers each containing two layers of convolution, batch normalization and a ReLu activation. Lastly a split policy/value head, where the output of the last residual layer is split into two different outputs. The policy head are two layers: first a final convolutional layer, and last a fully connected layer with a \textit{log softmax} layer. The value head, consists of three layers: first a convolutional layer, then a fully connected layer, and lastly a fully connected layer with a \textit{tanh} activation function. The Figure \ref{fig:nnarch}. depicts the architecture.

@@

\subsection{Convolutional Layers}

It is important to understand why we use convolutional layers when dealing with board games as convolution is often associated with image. This is because a convolutional layer is focused on merging multiple input parameters to a single neuron, making it an input parameter for the next layer representing the locality around the center point of the original input. Playing the game of Breakthrough, a piece in only able to capture pieces in its immediate vicinity. This lead to the selection of a convolutional kernel with the size of $3x3$ and a stride of $1$.

The knowledge of each piece moves isn't enough to represent the whole gamestate within the neural network. This is why we use multiple residual layers with convolution. These residual layers stack convolutions on each other making the final convolution represent the locality of all the other localities, allowing the neural network to have a representation of the whole gamestate in its parameters.

The selection of parameters was done for these reasons as well as to most closely resemble the architecture described in AlphaZero.

\subsection{Residual Layers}

The residual layers as a concept remember the output of the previous layer and add the results of the residual layer and the previous layer together. On a higher level this leads to the internal representation of the neural network to maintain the whole game boards in its representation, s.t. two areas that are far away from each other maintain the same level of locality as two that are close.

Why we use a residual layer for applying a convolution layer multiple times is to gain locality of the whole game board. A residual layer applies a function like this $res(x) = x + l(x)$ where $l(x)$ is the function of the layer, commonly a convolutional layer.

\subsection{Policy Head}

The policy head of the neural network returns a vector of the size of the action space $w * h * 6$ where $w$ is the width of the board $h$ is the height of the board and $6$ represents the six cardinal direction pawns can move (straight, and diagonally both left and right for the first player, and backwards for the second player). The vector is then masked s.t. the values representing moves that can not be played on the board are given the value of $0$.

\subsection{Value Head}

The value head of the neural network returns a single value representing the predicted value of the neural network. This predicted value is trained to be the end value of the game after taking the predicted move.

@@

The data collected is $(\pi, \tau, p, v)$, where $\pi$ is the action probabilities provided by 
MCTS for $N$ iterations, $\tau$ is the reward for the whole episode, $1$ if white wins,
$-1$ if black wins, $p$ is the policy vector
from the neural network, and $v$ is the predicted reward of the game by the neural network.

This data collection trains the neural network in such a way that it is a function approximation for a single state to return the values that would have been returned from applying MCTS for $N$ iterations.

@@

\section{Explainable state representations}

In this section we describe the

\subsection{Testing With Concept Activation Vectors}

Once the neural network has learned to play the game of Breakthrough, we examine its internal state w.r.t HLC's that we understand. The first examined HLC is numbers advantage, that is, the number of pieces that a player has over his opponent. The higher level idea to a human player would be that they are in a better position since they have more pieces.

To examine the higher level component we take a look at the internal state of the neural network itself. To do this we take a state, we run the state through the neural network, and while the state is propagating through the network we select a layer to split the network, we call that layer $l_{split}$. At that point in time the state is a mutated vector $l_{split}(l_{split-1}(...l_0(state)))$ of the initial state. We save that vector as a point in the N-dimensional space that it represents, and once we've gathered enough points in that N-dimensional space we're able to train a linear classifier using the HLC as a label. We train a Stochastic Gradient Descent classifier, from the SKLearn library, to construct a hyperplane that splits the space into two binary states, one that contains the HLC and another that doesn't. 
\begin{figure}[]
    \centering
    \includegraphics[width=0.7\textwidth]{graphics/linear_separation}
    \caption{Trained linear classifier on a 2-dimensional space}
    \label{fig:scattersplit}
\end{figure}
An example of how this would look with only two dimensions is shown in Figure \ref{fig:scattersplit} Once we've successfully trained this linear classifier we can run another state $s$ through the neural network. Once the prediction has finished, we view the selected action $a$ of $s$ from the policy vector $p$ from the neural network. We view the loss of that selected action $a$. Then we apply the backpropagation algorithm up to that same layer $l_{split}$ and evaluate the gradient there. If that gradient moves to the direction of the HLC we say that the state includes the HLC, and doesn't if it move away from the HLC.
\begin{figure}[]
    \centering
    \includegraphics[width=0.7\textwidth]{graphics/linear_separation_with_direction}
    \caption{Trained linear classifier on a 2-dimensional space with arrow representing gradient}
    \label{fig:scattersplitarrow}
\end{figure}
We show an example in Figure \ref{fig:scattersplitarrow}. where the arrow in the image represents the direction the gradient is moving, if it moves toward the HLC the state $s$ includes the HLC otherwise it does not. Using this method we evaluate the internal representation of the state within the neural network, and are able to see whether it recognizes the HLC.

### 2020-08-22 latex/sections/results.tex
latexing

\ifdraft{In this section you discuss any issues that came up while developing
the system.  If you found something particularly interesting,
difficult, or an important learning experience, put it here.  This is
also a good place to put additional figures and data.}
hello world

>>>>

\ifdraft{In this section you discuss any issues that came up while developing
the system.  If you found something particularly interesting,
difficult, or an important learning experience, put it here.  This is
also a good place to put additional figures and data.}

In this section we discuss the results of considering HLC within the game Breakthrough, using a Concept Activation vector for evaluating how the neural network recognizes the HLC's. We first take a look at the changes in ephasis of the neural network during training. The main point of interest there being whether the nerual network notices simple HLCs early then stops taking them into consideration as the network improves.

\section{Evaluating the improvements of the neural network over generations}

To test which HLC the neural network places its emphasis on during training we trained a neural network for $200$ iterations, taking snapshots of the network every $10$ generations. Then we had the neural network play against itself for $100$ games collecting the states it encountered during play. These states were then examined by a concept activation vector representing these HLCs. Firstly there is the numbers HLC, where the number of pawn the player has minus the number of pawns the opponent had, the breakpoint we selected was $2$ meaning that if you have $2$ more pawns then your opponent, you're in a position where a HLC called numbers advantage is present.

\begin{figure}[]
    \centering
    \includegraphics[width=0.7\textwidth]{graphics/number_pawns_trend}
    \caption{Percentage of selected states containing the HLC numbers advantage}
    \label{fig:numberadvantage}
\end{figure}

The Figure \ref{fig:numberadvantage}. shows that over the course of training the numbers advantage HLC is only ever a slight factor in the selection of states and we can say that the neural network doesn't really consider number advantage in its selection process.

The second HLC we examined was aggressiveness, that is a state in which you most advanced pawn is $2$ or more squares further than the opponents most advanced pawn.

\begin{figure}[]
    \centering
    \includegraphics[width=0.7\textwidth]{graphics/most_advanced_trend.png}
    \caption{Percentage of selected states containing the HLC aggressiveness}
    \label{fig:aggressiveness}
\end{figure}

From Figure \ref{fig:aggressiveness}. we can see that the aggressiveness HLC is a growing factor over as the nerual network is trained. Generally when your opponent is not skilled this strategy is considered a good one.

The last HLC that we examined was unity, the unity HLC represents the absolute average distance of your pawns from the center row of your pawns. This HLC is calculated as the row of your furthest pawn from the starting row $r_{far}$, the row of your nearest pawn from the starting row $r_{near}$. Finding the middle row is then $\frac{r_{far} + r_{near}}{2}$, we then take the absolute of the average distance from the pawns to that row. To find the point at which this value relates to a state with the HLC unity, we sampled a myriad of states and decided on the value of $0.35$. The value of $0.35$ generally allows your states to have two rows that have the majority of the pawns and one or two pawns one row away from the group.

\begin{figure}[]
    \centering
    \includegraphics[width=0.7\textwidth]{graphics/unity_trend.png}
    \caption{Percentage of selected states containing the HLC aggressiveness}
    \label{fig:unity}
\end{figure}

The literature of breakthrough implies that being patient and waiting until your opponent makes a mistake is generally the preferred strategy for winning, and the concept activation vector for Unity is generally triggered, and increases with the amount of generations the neural network is trained for. These results imply that the neural network does infact recognize successful strategies, and focuses its training in the direction of the successful strategies.
